\plaintitle

### 2016-12-22 paper/paper.tex
submitted version of paper

%% EXAMPLE BEGIN -- HOW TO OVERRIDE THE DEFAULT COPYRIGHT STRIP -- (July 22, 2013 - Paul Baumann)
% \toappear{Permission to make digital or hard copies of all or part of this work for personal or classroom use is      granted without fee provided that copies are not made or distributed for profit or commercial advantage and that copies bear this notice and the full citation on the first page. Copyrights for components of this work owned by others than ACM must be honored. Abstracting with credit is permitted. To copy otherwise, or republish, to post on servers or to redistribute to lists, requires prior specific permission and/or a fee. Request permissions from permissions@acm.org. \\
% {\emph{CHI'14}}, April 26--May 1, 2014, Toronto, Canada. \\
% Copyright \copyright~2014 ACM ISBN/14/04...\$15.00. \\
% DOI string from ACM form confirmation}
%% EXAMPLE END -- HOW TO OVERRIDE THE DEFAULT COPYRIGHT STRIP -- (July 22, 2013 - Paul Baumann)

@@

% Paper metadata (use plain text, for PDF inclusion and later
% re-using, if desired).  Use \emtpyauthor when submitting for review
% so you remain anonymous.
\def\plaintitle{Lifting the Curse of Dimensionality: High-Dimensional Visual Informatics using d3}
\def\plainauthor{Geoffrey G. Roeder}
%\def\emptyauthor{}
\def\plainkeywords{Authors' choice; of terms; separated; by
  semicolons; include commas, within terms only; required.}
\def\plaingeneralterms{Documentation, Standardization}

@@

\numberofauthors{3}
\author{%
  \alignauthor{Geoffrey G. Roeder\\
    \affaddr{University of Toronto}\\
    \email{roeder@cs.toronto.edu}}\\
}

@@

PLACEHOLDER To address this visualization gap between classical statistics and ML, my project will produce a new software tool that implements automated dimensionality reduction algorithms from the domain of visual analytics like t-SNE (Hinton \& Van der Maatens 2008), Multi-Dimensional Scaling (Kruskal \& Wish, 1978) and Sammon Mapping (Sammon, 1969) in the context of a D3.js visualization interface. The interface will provide users with multiple reduced dimensionality versions of a dataset that can be clustered using the algorithms DBSCAN density-based clustering, (Ester et al. 1996), k-means centroid-based clustering (Hartigan \& Wong 1979), or mixture-of-Gaussians distribution-based clustering. The D3.js interface is a way for the tool to present the visualizations in an easy-to-modify form. The software provides easy-to-use sliders for adjusting the parameters and viewing boxes that allow a researcher to experiment and iterate quickly over different visualizations 

@@

Classical statistics has some tools to offer, like a matrix-style pairwise scatterplot showing how each pair of feature changes with respect to every other, accessible through the \texttt{pairs} function in the R programming language. % 

@@

% TODO: maybe a figure here to motivate, as in the presentaiton

@@

% TODO: cite Tableau

@@

In contrast, the Visualization Dashboard encapsulates the statistical experitise needed to program an advanced algorithm by exposing only sliders and the algorithm output as a scatterplot. %

@@

There have been a number of recent research contributions that attempt to acheive a similar level of technical encapsulation for ML, and we review these in this section. %

@@

\footnote{A decision boundary is a region of the data space in which any point that falls in the region is classified identically. This is commonly visualized by colouring entire decision regions uniformly in a scatterplot of the points.}. %

@@

In the Setosa.io interface, users can click and drag points diplayed on a scatterplot to see how changing the location of inputs influences the projection onto 1 or 2 dimensions. %

@@

\section{Visualization Dashboard}
The Visualization Dashboard is a web application that runs locally in a user's browser. %
%
The central visualization component of the interface is shown in Figure 1. %
%
This section focuses on what a user is able to acheive with the interface, and the design choices that were made to facilitate this. %
%
\subsection{Technical Overview}
First, a brief description of the software used to develop the Visualization Dashboard is presented, intended to help other researchers select among the available libraries online. %
%
The highly developed state of open source ML Javascript libraries makes the development of experimental visualization software like the Visualization Dashboard much less resource intensive than it would have been in the recent past.%
%
\\\\
%
The dimensionality reduction algorithms were implemented using \texttt{ml-js} \cite{mljs} for PCA and general linear algebra, \texttt{tsnejs} \cite{tSNEJS} for tSNE, and \texttt{mdsjs} \cite{mdsjs} for the Multidimensional Scaling (MDS). %
%
Plotting functionality in 1D and 2D was implemented using D3.js \cite{d3js}, and 3D plotting was implemented using the Highcharts \cite{highcharts} framework. %
%
Automated data parsing was implemented using the \texttt{Papa Parse} \cite{papaparse} library.
%

@@

These choices are meant to draw attention away from programatic data loading issues and convey a sense of simplicity. %

@@

\footnote{The library handling data parsing, Papa Parse \cite{papaparse} is, robust to data formatting and will infer the format of the data from the first few lines (e.g. tab-separated, comma-separated, etc.). %

@@

Although the prototype of the Visualization Dashboard indicates that only comma-separated values will be accepted, \cite{papaparse} can easily be extended to arbitrary data formats meaning that a user will not need to pay attention to how the data is formatted.}. %

@@

Each dimensionality reduction widget like the one in Figure 1 initially contains three blank subdivisions that can be populated with scatterplots based on the reduced-dimensionality data (see Figure 1). %

@@

Scatterplots were chosen for as the key visualization component of the Visualization Dashboard because they are a model-agnostic way of viewing data in 1, 2, or 3 dimensions. %

@@

Hence, the index allows a user to investigate the points back in the original high-dimensional space to explore what correspondances might by causing their lower-dimensional distance. %

@@

The second, more speculative reason is to stoke the curiousity of a user about the changes that may occur next in a visualization as they vary parameters under some rationale. %

@@

The ML community, like many other subdisciplines of the field of Comptuer Science, is strongly connected online. %

@@

For example, next to the t-SNE algorithm is a hyperlink to the recent Google Brain project \cite{wattenberg2016how} that explores how to tune the t-SNE algorithm when the true high-dimensional dataset has a particular structure. %

@@

By playing around with the parameters of t-SNE\footnote{To reproduce this result, use set LR to 100, P to 50, and N to 500.}, %

@@

This shows that t-SNE discovers a lower dimensional represntation that ignores the curvature of the embedded manifold in 3D and represents the high-dimensional structure correctly, using only 2 dimensions. %

@@

The participant was informed that the interface was useful for ML practitioners to understand high dimensional data. %
%
Without further instruction, the participant was then shown the interface, and asked to explore its functionality. %
%
The participant inferred that a data file could be dropped into the application, and that the interface would react to the data file by showing different visualizations of it. %
%
This suggests that no major flaws exist in the design concept. %
%
\\\\
%
The participant was then given access to a data file and asked to use the interface to explore it. %
%
After loading the file in the interface, he successfully generated a number of different visualizations. %
%
One behaviour of note is that after the three-dimensional visualization was discovered, which allows rotation around the point cloud, the participant attempted to click and dragging the two and one dimensional visualizations in order to elicit a similar response from the view. %%
%
When asked about the behaviour, the participant commented that the ability to pan or zoom is necessary to understand a dataset under exploration. %
%
\\\\
%
A final noteworthy result came from a request by the participant who asked for a tabular view of the raw data itself that could be modified with extra synthetic data. %
%
The paricipant noted that this would be useful for understanding intuitively how the reduced-dimensional representations are formed. %
%
The formative study suggests that the interface is effective in allowing exploration and study of complex datasets. %
%
The kinds of visualizations and play allowed by the current interface could be fruitfully extended in future enhancements to the interface according to the suggestions given. %
%
These enhancements can be built into the existing framework in a straightforward manner, now that the core functionality has been developed. %
%

\section{Discussion and Future Work}%
%
\begin{figure*}
  \centering
  \includegraphics[width=1.5\columnwidth]{figures/extended}
  \caption{An extended Data Space Model that captures exploratory data visualizations}~\label{fig:figure2}
\end{figure*}
%
Based on the formative study, the author concludes that the interface is ready for use as a research platform into high-dimensional visualization.%
%
Moreover, the process of producing this interface has provided an opportunity to consider how current theoretical perspectives in the field of information visualization can be leveraged to understand the Visualization Dashboard. %
%
Chi's Data Space Model of information visualization is useful here as a model of the user interacting with the Visualization Dashboard.%
%
Chi's model is designed to capture the process of analysis that yields multiple visualization of a dataset. %
%
Raw data is transformed through four stages: first to an analytical abstraction representing the data as some abstract data structure like a graph, tree, or vector, next to a visualization abstraction like a network diagram, point set, or surface, and finally to a view of the data that displays some salient features of interest. %
%
\\\\
%
Chi's model is an extension of Stuart Card's Information Visualization Pipeline that imagines an analyst proceding through a single path from raw data to final view \cite{chi1999framework}. %
%
Chi's model extends Card's to allow multiple paths, some of which terminate before yielding a final visualization. %
%
The Visualization Dashboard motivates an extension to Chi's model that can capture the process by which a ML researcher learns about data using multiple, temporary visualizations. %
%
The extension involves the addition of backlinks from the visual abstraction layer in Chi's model to previous layers, effectively permitting cycles through the network (see Figure 5 for a simple example). %
%
The backlinks represent "discovery links", where the current visual abstraction has given some useful information that motivates a new brach from an earlier node in the path.
%
Such an extended Data Space Model tracks the heuristic steps that an analyst takes in discovering relevant structure by tracing the links among the diagrams. %
%
This model could serve as a map for other researchers looking at similarly data to help guide their exploration of the high dimensional space, or simply provide a slightly richer framework for analyzing visualization systems. %
%
The scope of the present paper does not include a strong theoretical aims, so a further development of this contribution will be left to future projects. %
%
\section{Conclusion}%
This paper has introduced the Visualization Dashboard, a high-dimensional data exploration tool designed to support ML researchers in the early phase of model building. Design choices have been described and explored in the context of the intended use of the tool. A formative study has shown that the tool is ready to be used as a platform to answer research questions, and a novel extension to the Data Space Model has been proposed to describe the use of visualizations by a user of the Visualization Dashboard.
%

>>>>

%% EXAMPLE BEGIN -- HOW TO OVERRIDE THE DEFAULT COPYRIGHT STRIP -- (July 22, 2013 - Paul Baumann)
\makeatletter
\def\@copyrightspace{\relax}
\makeatother
%% EXAMPLE END -- HOW TO OVERRIDE THE DEFAULT COPYRIGHT STRIP -- (July 22, 2013 - Paul Baumann)

@@

% Paper metadata (use plain text, for PDF inclusion and later
% re-using, if desired).  Use \emtpyauthor when submitting for review
% so you remain anonymous.
\def\plaintitle{The Visualization Dashboard: A Browser-Based Tool for High-Dimensional Data Exploration}
\def\plainauthor{Geoffrey Roeder}
%\def\emptyauthor{}
\def\plainkeywords{Authors' choice; of terms; separated; by
  semicolons; include commas, within terms only; required.}
\def\plaingeneralterms{Documentation, Standardization}

@@

\numberofauthors{3}
\author{%
  \alignauthor{Geoffrey Roeder\\
    \affaddr{University of Toronto}\\
    \email{roeder@cs.toronto.edu}}\\
}

@@

This research addresses a visualization gap between the needs of classical statistics and current practice in machine learning by introducing the Visualization Dashboard, a new tool for high dimensional data exploration. %
The Dashboard provides users with multiple state-of-the-art dimensionality reduction and clustering algorithms for exploring high-dimensional data. %
The Dashboard is designed for use in early phases of statistical and machine learning analyses, to assist in comprehending data for model exploration and feature extraction purposes. %
A review of prior work in this area shows it has focused primarily on synthetic data for pedagogical aims. %
Some prior work, hosted online, is integrated into the Dashboard to enrich user comprehension of parameter tuning in dimensionality reduction. %
Motivating design choices and functionality of the Dashboard are discussed in detail, followed by the results of a formative study and implications for information visualization theory.

@@

Classical statistics has some tools to offer, like a matrix-style pairwise scatter plot showing how each pair of feature changes with respect to every other, accessible through the \texttt{pairs} function in the R programming language. % 

@@

@@

@@

In contrast, the Visualization Dashboard encapsulates the statistical expertise needed to program an advanced algorithm by exposing only sliders and the algorithm output as a scatter plot. %

@@

There have been a number of recent research contributions that attempt to achieve a similar level of technical encapsulation for ML, and we review these in this section. %

@@

\footnote{A decision boundary is a region of the data space in which any point that falls in the region is classified identically. This is commonly visualized by colouring entire decision regions uniformly in a scatter plot of the points.}. %

@@

In the Setosa.io interface, users can click and drag points displayed on a scatter plot to see how changing the location of inputs influences the projection onto 1 or 2 dimensions. %

@@

\section{Visualization Dashboard}
The Visualization Dashboard is a web application that runs locally in a user's browser. %
%
The central visualization component of the interface is shown in Figure 1. %
%
This section focuses on what a user is able to achieve with the interface, and the design choices that were made to facilitate this. %
%
\subsection{Technical Overview}
First, a brief description of the software used to develop the Visualization Dashboard is presented, intended to help other researchers select among the available libraries online. %
%
The highly developed state of open source ML Javascript libraries makes the development of experimental visualization software like the Visualization Dashboard much less resource intensive than it would have been in the recent past.%
%
\\\\
%
The dimensionality reduction algorithms were implemented using \texttt{ml-js} \cite{mljs} for PCA and general linear algebra, \texttt{tsnejs} \cite{tSNEJS} for tSNE, and \texttt{mdsjs} \cite{mdsjs} for the Multidimensional Scaling (MDS). %
%
Plotting functionality in 1D and 2D was implemented using D3.js \cite{d3js}, and 3D plotting was implemented using the Highcharts \cite{highcharts} framework. %
%
Automated data parsing was implemented using the \texttt{Papa Parse} \cite{papaparse} library.
%

@@

These choices are meant to draw attention away from programmatic data loading issues and convey a sense of simplicity. %

@@

\footnote{The library chosen for data parsing, Papa Parse \cite{papaparse} can infer the format of the data from the first few lines (e.g. tab-separated, comma-separated, etc.). %

@@

Although the prototype of the Visualization Dashboard recommends only comma-separated values, \cite{papaparse} can easily be extended to arbitrary data formats meaning that a user will not need to pay attention to how the data is formatted.}. %

@@

Each dimensionality reduction widget like the one in Figure 1 initially contains three blank subdivisions that can be populated with scatter plots based on the reduced-dimensionality data (see Figure 1). %

@@

Scatter plots were chosen for as the key visualization component of the Visualization Dashboard because they are a model-agnostic way of viewing data in 1, 2, or 3 dimensions. %

@@

Hence, the index allows a user to investigate the points back in the original high-dimensional space to explore what correspondences might by causing their lower-dimensional distance. %

@@

The second, more speculative reason is to stoke the curiosity of a user about the changes that may occur next in a visualization as they vary parameters under some rationale. %

@@

The ML community, like many other subdisciplines of the field of Computer Science, is strongly connected online. %

@@

For example, next to the tSNE algorithm is a hyperlink to the recent Google Brain project \cite{wattenberg2016how} that explores how to tune the tSNE algorithm when the true high-dimensional dataset has a particular structure. %

@@

By playing around with the parameters of tSNE\footnote{To reproduce this result with the dataset swiss\_roll\_pi.csv (email author for access), set LR to 100, P to 50, and N to 500.}, %

@@

This shows that tSNE discovers a lower dimensional representation that ignores the curvature of the embedded manifold in 3D and represents the high-dimensional structure correctly, using only 2 dimensions. %

@@

The participant was informed that the interface was useful for ML practitioners to understand high dimensional data. %
Without further instruction, the participant was then shown the interface, and asked to explore its functionality. %
The participant inferred that a data file could be dropped into the application, and that the interface would react to the data file by showing different visualizations of it. %
This suggests that no major flaws exist in the design concept. %
The participant was then given access to a data file and asked to use the interface to explore it. %
After loading the file in the interface, he successfully generated a number of different visualizations. %
One behaviour of note is that after the three-dimensional visualization was discovered, which allows rotation around the point cloud, the participant attempted to click and dragging the two and one dimensional visualizations in order to elicit a similar response from the view. %%
When asked about the behaviour, the participant commented that the ability to pan or zoom is necessary to understand a dataset under exploration. %
A final noteworthy result came from a request by the participant who asked for a tabular view of the raw data itself that could be modified with extra synthetic data. %
The paricipant noted that this would be useful for understanding intuitively how the reduced-dimensional representations are formed. %
The formative study suggests that the interface is effective in allowing exploration and study of complex datasets. %
The kinds of visualizations and play allowed by the current interface could be fruitfully extended in future enhancements to the interface according to the suggestions given. %
These enhancements can be built into the existing framework in a straightforward manner, now that the core functionality has been developed. %
\section{Discussion and Future Work}%
\begin{figure*}
  \includegraphics[width=2\columnwidth]{figures/extended}
  \caption{An extended Data Space Model that captures exploratory data visualizations}~\label{fig:figure2}
\end{figure*}
Based on the formative study, the author concludes that the interface is ready for use as a research platform into high-dimensional visualization.%
Moreover, the process of producing this interface has provided an opportunity to consider how current theoretical perspectives in the field of information visualization can be leveraged to understand the Visualization Dashboard. %
Chi's Data Space Model of information visualization is useful here as a model of the user interacting with the Visualization Dashboard.%
Chi's model is designed to capture the process of analysis that yields multiple visualization of a dataset. %
Raw data is transformed through four stages: first to an analytical abstraction representing the data as some abstract data structure like a graph, tree, or vector, next to a visualization abstraction like a network diagram, point set, or surface, and finally to a view of the data that displays some salient features of interest. %
Chi's model is an extension of Stuart Card's Information Visualization Pipeline that imagines an analyst proceeding through a single path from raw data to final view \cite{chi1999framework}. %
Chi's model extends Card's to allow multiple paths, some of which terminate before yielding a final visualization. %
The Visualization Dashboard motivates an extension to Chi's model that can capture the process by which a ML researcher learns about data using multiple, temporary visualizations. %
The extension involves the addition of backlinks from the visual abstraction layer in Chi's model to previous layers, effectively permitting cycles through the network (see Figure 5 for a simple example). %
The backlinks represent "discovery links", where the current visual abstraction has given some useful information that motivates a new branch from an earlier node in the path.
Such an extended Data Space Model tracks the heuristic steps that an analyst takes in discovering relevant structure by tracing the links among the diagrams. %
This model could serve as a map for other researchers looking at similarly data to help guide their exploration of the high dimensional space, or simply provide a slightly richer framework for analyzing visualization systems. %
The scope of the present paper does not include a strong theoretical aims, so a further development of this contribution will be left to future projects. %
\section{Conclusion}%
This paper has introduced the Visualization Dashboard, a high-dimensional data exploration tool designed to support ML researchers in the early phase of model building. Design choices have been described and explored in the context of the intended use of the tool. A formative study has shown that the tool is ready to be used as a platform to answer research questions, and a novel extension to the Data Space Model has been proposed to describe the use of visualizations by a user of the Visualization Dashboard.
